Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2025-01-16 ms-3/ms.tex
Continue finalizing manuscript

\clearpage
\section*{Introduction}
Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
Short-term fluctuations in incidence time series are often caused in part by variable reporting, but also reflect the population-level impacts of transmission heterogeneity, or changes in susceptiblility~\citep{lloyd2005superspreading, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.

Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers, quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}, indicating temporal changes in the level of variability around the expected number of secondary infections that will result if a particular individual becomes infected.
However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.
In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability~\citep{kretzschmar2020impact}, there is a need for complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies was required for SARS-CoV-2 control~\citep{sun2021transmission}. 
An important challenge therefore is to develop methods that can detect changes in population-level variability in incidence time series, and to interpret these changes in terms of underlying transmission processes.

Emerging statistical techniques are leveraging variability in epidemic time series to enhance understanding of disease dynamics at the population level. 
For example, a recently-developed method estimates the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual, directly from population-level incidence data~\citep{kirkegaard2021superspreading}. 
Clustering of cases has also been estimated directly from incidence data~\citep{schneckenreither2023assessing}.
Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\citep{graham2019measles}. 
Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with relevance for the sensitivity of cities to forcing from exogenous drivers of transmission (e.g., climate) and for surge capacity in the health system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.

What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
Under a wide range of configurations for a branching process model of contagion spread the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter. 
The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$. 
Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
The negative binomial distribution may also accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial. 
Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion~\citep{barron1992analysis}.

@@

Since very large $\theta$ correspond to processes that are close to a Poisson process, the test will not produce a p-value if any of the three values are greater than 1/(1e-3). 
Very large $\theta$ estimates are often produced in small populations when the process can't be distinguished from a Poisson process. Therefore, we have very small $\phi = 1/\theta$  values be considered zero, at which point we fail to reject the Poisson hypothesis.

>>>>

\clearpage
\section*{Introduction}
Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
Short-term fluctuations in incidence time series are caused in part by variable reporting, but may also reflect the population-level impacts of transmission heterogeneity, and changes in the landscape of susceptiblility~\citep{lloyd2005superspreading,lloyd2007maximum, lau2017spatial, dalziel2018urbanization, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.

Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers, quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}. 
Similar evidence has been recovered through statistical reconstruction of transmission networks, indicating temporal trends in the level of dispersion at different phases of an epidemic~\citep{lau2017spatial}.
However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.
In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and scalability~\citep{kretzschmar2020impact}, it would be helpful to have complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies were required for SARS-CoV-2 control during the early phases of the pandemic in China~\citep{sun2021transmission}. 
An important challenge therefore is to develop methods that can detect changes in population-level variability in incidence time series, and to interpret these changes in terms of underlying transmission processes.

Emerging statistical techniques are leveraging variability in epidemic time series to enhance understanding of disease dynamics at the population level. 
For example, a recently-developed method uses population-level incidence data to the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual~\citep{kirkegaard2021superspreading}. 
It is also possible to estimate the dispersion parameter from the distribution of the final size of a series of localized outbreaks~\citep{blumberg2013inference}.
Clustering of cases has also been estimated directly from incidence data~\citep{schneckenreither2023assessing}.
Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\citep{graham2019measles}. 
Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with implications on the sensitivity of cities to climate forcing and for surge capacity in the health system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.

What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
Under a wide range of configurations for a branching process model of contagion spread, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter. 
The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$. 
Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
The negative binomial distribution may also accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial. 
Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion~\citep{barron1992analysis}.

@@

Since very large $\theta$ correspond to processes that are operationally nearly identical to a Poisson process, the test will not produce a p-value if any of the three values are greater than a user-specified threshold. 
In the applicatio below we set this threshold at $10^3$, meaning that theta estimates with temporal crowding that was within $0.1\%$ of that expected for a Poisson process was considered effectively Poisson.

Very large $\theta$ estimates are often produced in small populations when the process can't be distinguished from a Poisson process. Therefore, we have very small $\phi = 1/\theta$  values be considered zero, at which point we fail to reject the Poisson hypothesis.
